---
title: "ATAC-seq differential accessibility in mouse hippocampus"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6.5,
  dpi = 150,
  fig.align = "center"
)
```

## Data import and sample annotation

```{r libraries}
library(SummarizedExperiment)
library(GenomicRanges)
library(edgeR)
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(dplyr)
library(tibble)
```

```{r import-data}
se <- readRDS("mouse_mm38_hippocampus.peakCounts.SE.rds")
se
```

The object contains one assay of ATAC-seq peak counts and 12 samples. Since the original `colData` only contains sequencing depth, stress status and sex were reconstructed from the sample names.

```{r sample-annotation}
build_sample_df <- function(se) {
  samples <- colnames(se)
  sample_df <- data.frame(
    sample = samples,
    stress = ifelse(grepl("^FSS_", samples), "FSS", "Control"),
    sex = ifelse(grepl("_female_", samples), "female", "male"),
    depth = colData(se)$depth,
    row.names = samples
  )
  sample_df$stress <- factor(sample_df$stress, levels = c("Control", "FSS"))
  sample_df$sex <- factor(sample_df$sex, levels = c("female", "male"))
  sample_df
}

sample_df <- build_sample_df(se)
colData(se) <- cbind(colData(se), DataFrame(sample_df))

table(colData(se)$stress, colData(se)$sex)
```

## Exploratory QC

TMM-normalized log-CPM values were used for exploratory sample-level QC.

```{r qc-objects}
y <- DGEList(counts = assay(se, "counts"))
y <- calcNormFactors(y, method = "TMM")
logcpm <- cpm(y, log = TRUE, prior.count = 2)

ann <- data.frame(
  stress = sample_df$stress,
  sex = sample_df$sex,
  row.names = rownames(sample_df)
)

ann_colors <- list(
  stress = c(Control = "#4DBBD5", FSS = "#E64B35"),
  sex = c(female = "#D8B70A", male = "#00A087")
)

display_names <- sub("\\.bam$", "", colnames(se))
```

```{r corr-heatmap, fig.width=11, fig.height=9}
cor_mat <- cor(logcpm)

pheatmap(
  cor_mat,
  labels_row = display_names,
  labels_col = display_names,
  annotation_row = ann,
  annotation_col = ann,
  annotation_colors = ann_colors,
  fontsize = 9,
  fontsize_row = 8,
  fontsize_col = 8,
  angle_col = 45,
  border_color = NA,
  main = "Sample-sample correlation (TMM logCPM)"
)
```

```{r mds-plot, fig.width=10, fig.height=7}
mds <- plotMDS(y, plot = FALSE)
mds_df <- data.frame(
  x = mds$x,
  y = mds$y,
  sample = display_names,
  stress = sample_df$stress,
  sex = sample_df$sex,
  depth = sample_df$depth
)

ggplot(mds_df, aes(x = x, y = y, color = sex, shape = stress, label = sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3.5, max.overlaps = Inf) +
  labs(
    title = "MDS plot of ATAC-seq samples",
    x = "Leading logFC dim 1",
    y = "Leading logFC dim 2"
  ) +
  theme_bw()
```

Two samples appear somewhat distinct, `Control_female_GSM6041843` and `FSS_female_GSM6041854`. Overall, separation by sex appears stronger than separation by stress condition.

```{r depth-vs-mds}
center_x <- mean(mds_df$x)
center_y <- mean(mds_df$y)
mds_df$dist_to_center <- sqrt((mds_df$x - center_x)^2 + (mds_df$y - center_y)^2)

cor.test(mds_df$dist_to_center, mds_df$depth, method = "spearman")

ggplot(mds_df, aes(x = dist_to_center, y = depth, label = sample, shape = stress)) +
  geom_point(aes(color = sex), size = 3) +
  geom_text_repel(size = 3.5, max.overlaps = Inf) +
  labs(
    title = "Distance to MDS centroid vs sequencing depth",
    x = "Distance to centroid in 2D MDS",
    y = "Sequencing depth"
  ) +
  theme_bw()
```

Sequencing depth may show a weak trend with distance to the MDS centroid, but this association is not statistically significant.

## Differential accessibility analysis

Differential accessibility was tested with edgeR quasi-likelihood GLMs using TMM normalization. For the stress analysis, sex was included as a covariate. For the sex analysis, stress was included as a covariate.

```{r analysis-functions}
prepare_dge <- function(se) {
  y <- DGEList(counts = assay(se, "counts"))
  calcNormFactors(y, method = "TMM")
}

run_da <- function(se, formula, coef_name = NULL, contrast = NULL) {
  sample_df <- build_sample_df(se)
  y <- prepare_dge(se)
  design <- model.matrix(formula, data = sample_df)

  y <- estimateDisp(y, design)
  fit <- glmQLFit(y, design)

  if (!is.null(coef_name)) {
    qlf <- glmQLFTest(fit, coef = coef_name)
    test_name <- coef_name
  } else if (!is.null(contrast)) {
    qlf <- glmQLFTest(fit, contrast = contrast)
    test_name <- "custom_contrast"
  } else {
    stop("Provide either coef_name or contrast.")
  }

  tab <- topTags(qlf, n = Inf)$table
  rr <- rowRanges(se)

  res <- tab %>%
    rownames_to_column("row_id") %>%
    mutate(
      idx = as.integer(row_id),
      chr = as.character(seqnames(rr))[idx],
      start = start(rr)[idx],
      end = end(rr)[idx],
      region = paste0(chr, ":", start, "-", end)
    ) %>%
    relocate(region, chr, start, end, row_id)

  list(
    sample_df = sample_df,
    y = y,
    design = design,
    fit = fit,
    qlf = qlf,
    results = res,
    test_name = test_name
  )
}

plot_pvalue_hist <- function(res, title = "P-value distribution") {
  ggplot(res, aes(x = PValue)) +
    geom_histogram(
      breaks = seq(0, 1, by = 0.05),
      fill = "grey70",
      color = "white"
    ) +
    scale_x_continuous(breaks = seq(0, 1, by = 0.1), limits = c(0, 1)) +
    labs(
      title = title,
      x = "P-value",
      y = "Number of regions"
    ) +
    theme_bw(base_size = 12)
}

plot_volcano <- function(
  res,
  title = "Volcano plot",
  p_cutoff = 0.05,
  fdr_cutoff = 0.05,
  lfc_cutoff = 1,
  n_label = 5
) {
  res2 <- res %>%
    mutate(
      neglog10P = -log10(PValue),
      status = case_when(
        FDR < fdr_cutoff & logFC >=  lfc_cutoff ~ "FDR < 0.05 & logFC >= cutoff",
        FDR < fdr_cutoff & logFC <= -lfc_cutoff ~ "FDR < 0.05 & logFC <= -cutoff",
        PValue < p_cutoff & logFC >=  lfc_cutoff ~ "P < 0.05 & logFC >= cutoff",
        PValue < p_cutoff & logFC <= -lfc_cutoff ~ "P < 0.05 & logFC <= -cutoff",
        TRUE ~ "NS"
      )
    )

  n_lab <- min(n_label, nrow(res2))

  top_lab <- res2 %>%
    arrange(FDR, PValue) %>%
    slice_head(n = n_lab) %>%
    mutate(label = region)

  ggplot(res2, aes(x = logFC, y = neglog10P)) +
    geom_point(aes(color = status), alpha = 0.8, size = 1.8) +
    geom_text_repel(
      data = top_lab,
      aes(label = label),
      size = 3.2,
      max.overlaps = Inf,
      box.padding = 0.35,
      point.padding = 0.2,
      segment.linewidth = 0.3
    ) +
    geom_vline(xintercept = c(-lfc_cutoff, lfc_cutoff), linetype = 2, linewidth = 0.4) +
    geom_hline(yintercept = -log10(p_cutoff), linetype = 2, linewidth = 0.4) +
    scale_color_manual(
      values = c(
        "NS" = "grey75",
        "P < 0.05 & logFC >= cutoff" = "#E64B35",
        "P < 0.05 & logFC <= -cutoff" = "#4DBBD5",
        "FDR < 0.05 & logFC >= cutoff" = "#B2182B",
        "FDR < 0.05 & logFC <= -cutoff" = "#2166AC"
      )
    ) +
    labs(
      title = title,
      x = "log2 fold-change",
      y = expression(-log[10](PValue)),
      color = NULL
    ) +
    theme_bw(base_size = 12) +
    theme(
      legend.position = "bottom",
      plot.title = element_text(face = "bold")
    )
}

plot_sig_heatmap <- function(
  se,
  da,
  which = c("pvalue", "fdr"),
  cutoff = 0.05,
  max_regions = 100,
  scale_rows = TRUE,
  title = NULL
) {
  which <- match.arg(which)

  sample_df <- da$sample_df
  y <- da$y
  res <- da$results

  if (which == "pvalue") {
    sel <- res %>% filter(PValue < cutoff) %>% arrange(PValue)
    default_title <- paste0("Regions with PValue < ", cutoff)
  } else {
    sel <- res %>% filter(FDR < cutoff) %>% arrange(FDR, PValue)
    default_title <- paste0("Regions with FDR < ", cutoff)
  }

  if (nrow(sel) == 0) {
    message("No regions passed the selected threshold.")
    return(invisible(NULL))
  }

  sel <- sel %>% slice_head(n = min(max_regions, nrow(sel)))

  mat <- cpm(y, log = TRUE, prior.count = 2)[sel$row_id, , drop = FALSE]
  rownames(mat) <- sel$region

  if (scale_rows) {
    mat <- t(scale(t(mat)))
    mat[is.na(mat)] <- 0
  }

  ann <- data.frame(
    stress = sample_df$stress,
    sex = sample_df$sex,
    depth = sample_df$depth,
    row.names = rownames(sample_df)
  )
  ann <- ann[colnames(mat), , drop = FALSE]

  pheatmap(
    mat,
    annotation_col = ann,
    labels_col = sub("\\.bam$", "", colnames(mat)),
    show_rownames = nrow(mat) <= 40,
    fontsize_col = 8,
    fontsize_row = 7,
    angle_col = 45,
    border_color = NA,
    main = if (is.null(title)) default_title else title
  )
}

summarize_da <- function(da, fdr_thr = 0.05) {
  res <- da$results
  best <- res[which.min(res$FDR), ]
  data.frame(
    test = da$test_name,
    n_regions = nrow(res),
    n_fdr_sig = sum(res$FDR < fdr_thr, na.rm = TRUE),
    top_region = best$region,
    top_fdr = best$FDR
  )
}

run_and_plot_da <- function(se, formula, coef_name, prefix, max_regions = 100) {
  da <- run_da(se, formula = formula, coef_name = coef_name)

  print(summarize_da(da))
  print(head(da$results[, c("region", "logFC", "logCPM", "PValue", "FDR")], 10))
  print(plot_pvalue_hist(da$results, paste(prefix, "p-value distribution")))
  print(plot_volcano(da$results, title = paste(prefix, "volcano plot")))

  plot_sig_heatmap(
    se, da,
    which = "pvalue",
    cutoff = 0.05,
    max_regions = max_regions,
    title = paste(prefix, ": regions with PValue < 0.05")
  )

  plot_sig_heatmap(
    se, da,
    which = "fdr",
    cutoff = 0.05,
    max_regions = max_regions,
    title = paste(prefix, ": regions with FDR < 0.05")
  )

  invisible(da)
}

drop_samples <- function(se, samples_to_drop) {
  se[, !colnames(se) %in% samples_to_drop]
}

compare_models <- function(...) {
  das <- list(...)
  nm <- names(das)
  bind_rows(lapply(seq_along(das), function(i) {
    da <- das[[i]]
    res <- da$results
    data.frame(
      model = if (length(nm) && nzchar(nm[i])) nm[i] else paste0("model_", i),
      test = da$test_name,
      n_regions = nrow(res),
      n_pval_lt_0_05 = sum(res$PValue < 0.05, na.rm = TRUE),
      n_fdr_lt_0_05 = sum(res$FDR < 0.05, na.rm = TRUE),
      min_pvalue = min(res$PValue, na.rm = TRUE),
      min_fdr = min(res$FDR, na.rm = TRUE)
    )
  }))
}
```

### Stress vs control

```{r stress-full, fig.width=10, fig.height=9}
stress_da <- run_and_plot_da(
  se,
  formula = ~ sex + stress,
  coef_name = "stressFSS",
  prefix = "Stress vs Control"
)
```

The stress comparison did not yield FDR-significant regions at 0.05, although several regions showed low nominal p-values.

### Male vs female

```{r sex-full, fig.width=10, fig.height=9}
sex_da <- run_and_plot_da(
  se,
  formula = ~ stress + sex,
  coef_name = "sexmale",
  prefix = "Male vs Female"
)
```

The sex comparison showed a much stronger signal, with multiple regions remaining significant after FDR correction.

## Interaction model and sensitivity analysis

Given the stronger separation by sex in the exploratory QC, I also explored an interaction model to test whether the stress effect differed by sex. In addition, I repeated the analyses after excluding the two visually more isolated samples (`Control_female_GSM6041843.bam` and `FSS_female_GSM6041854.bam`) as a sensitivity analysis.

```{r filtered-se}
outlier_samples <- c(
  "Control_female_GSM6041843.bam",
  "FSS_female_GSM6041854.bam"
)

se_filt <- drop_samples(se, outlier_samples)
```

### Interaction model on full data

```{r interaction-full, fig.width=10, fig.height=9}
interaction_da <- run_and_plot_da(
  se,
  formula = ~ sex * stress,
  coef_name = "sexmale:stressFSS",
  prefix = "Interaction: sex × stress"
)
```

### Stress model after excluding two samples

```{r stress-filtered, fig.width=10, fig.height=9}
stress_da_filt <- run_and_plot_da(
  se_filt,
  formula = ~ sex + stress,
  coef_name = "stressFSS",
  prefix = "Stress vs Control (filtered)"
)
```

### Sex model after excluding two samples

```{r sex-filtered, fig.width=10, fig.height=9}
sex_da_filt <- run_and_plot_da(
  se_filt,
  formula = ~ stress + sex,
  coef_name = "sexmale",
  prefix = "Male vs Female (filtered)"
)
```

### Interaction model after excluding two samples

```{r interaction-filtered, fig.width=10, fig.height=9}
interaction_da_filt <- run_and_plot_da(
  se_filt,
  formula = ~ sex * stress,
  coef_name = "sexmale:stressFSS",
  prefix = "Interaction: sex × stress (filtered)"
)
```

## Summary table across models

```{r model-comparison-table}
model_summary <- compare_models(
  stress_full = stress_da,
  sex_full = sex_da,
  interaction_full = interaction_da,
  stress_filtered = stress_da_filt,
  sex_filtered = sex_da_filt,
  interaction_filtered = interaction_da_filt
)

model_summary
```

# Add depth as covariate 

```{r}
stress_depth_da <- run_and_plot_da(
  se,
  formula = ~ sex + depth + stress,
  coef_name = "stressFSS",
  prefix = "Stress adjusted for depth"
)

```

```{r}

sex_depth_da <- run_and_plot_da(
  se,
  formula = ~ stress + depth + sex,
  coef_name = "sexmale",
  prefix = "Sex adjusted for depth"
)
```

# Adding depth as covariate makes doesn't improve the analysis



