---
title: "Week 04 Assignment – p300, H3K27ac and H3K27me3 in mESC vs Heart"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(ensembldb)
  library(GenomicRanges)
  library(rtracklayer)
  library(epiwraps)
  library(ggplot2)
  library(ComplexHeatmap)
  library(GenomeInfoDb)
})

ah <- AnnotationHub::AnnotationHub(
  cache = "/mnt/auxiliary/.cache/R/AnnotationHub"
)

# Mouse GRCm38 EnsDb (Ensembl v102) – same build used in assignment_02
ensdb <- ah[["AH89211"]]   # EnsDb.Mmusculus.v102 / GRCm38

dir.create("peaks",  showWarnings = FALSE)
dir.create("tracks", showWarnings = FALSE)
```

---

# 1 – Downloading data from ENCODE

```{r download_peaks}
# ── mESC peaks ──────────────────────────────────────────────────────────────
# p300  – ENCSR000EYH  (ENCODE3, Bruce4 mESC, mm10)
download.file(
  "https://www.encodeproject.org/files/ENCFF653YVR/@@download/ENCFF653YVR.bed.gz",
  "peaks/mesc_p300.bed.gz", mode = "wb"
)

# H3K27ac – ENCSR000CGS (Bruce4 mESC, mm10, conservative IDR)
download.file(
  "https://www.encodeproject.org/files/ENCFF212HAG/@@download/ENCFF212HAG.bed.gz",
  "peaks/mesc_H3K27ac.bed.gz", mode = "wb"
)

# H3K27me3 – ENCSR000CGU (Bruce4 mESC, mm10, conservative IDR)
download.file(
  "https://www.encodeproject.org/files/ENCFF105NKG/@@download/ENCFF105NKG.bed.gz",
  "peaks/mesc_H3K27me3.bed.gz", mode = "wb"
)

# ── Heart peaks ──────────────────────────────────────────────────────────────
# p300  – ENCSR000EYF (adult heart, mm10)
download.file(
  "https://www.encodeproject.org/files/ENCFF097ULJ/@@download/ENCFF097ULJ.bed.gz",
  "peaks/heart_p300.bed.gz", mode = "wb"
)

# H3K27ac – ENCSR000CGR (adult heart, mm10, conservative IDR)
download.file(
  "https://www.encodeproject.org/files/ENCFF723ZZS/@@download/ENCFF723ZZS.bed.gz",
  "peaks/heart_H3K27ac.bed.gz", mode = "wb"
)

# H3K27me3 – ENCSR000CGQ (adult heart, mm10, conservative IDR)
download.file(
  "https://www.encodeproject.org/files/ENCFF183KJW/@@download/ENCFF183KJW.bed.gz",
  "peaks/heart_H3K27me3.bed.gz", mode = "wb"
)
```

```{r download_bigwigs}
# ── mESC bigwigs (fold change over control) ──────────────────────────────────
download.file(
  "https://www.encodeproject.org/files/ENCFF552AJY/@@download/ENCFF552AJY.bigWig",
  "tracks/mesc_p300.bw", mode = "wb"
)
download.file(
  "https://www.encodeproject.org/files/ENCFF610BPO/@@download/ENCFF610BPO.bigWig",
  "tracks/mesc_H3K27ac.bw", mode = "wb"
)
download.file(
  "https://www.encodeproject.org/files/ENCFF182FTP/@@download/ENCFF182FTP.bigWig",
  "tracks/mesc_H3K27me3.bw", mode = "wb"
)

# ── Heart bigwigs ─────────────────────────────────────────────────────────────
download.file(
  "https://www.encodeproject.org/files/ENCFF460PUH/@@download/ENCFF460PUH.bigWig",
  "tracks/heart_p300.bw", mode = "wb"
)
download.file(
  "https://www.encodeproject.org/files/ENCFF266FCV/@@download/ENCFF266FCV.bigWig",
  "tracks/heart_H3K27ac.bw", mode = "wb"
)
download.file(
  "https://www.encodeproject.org/files/ENCFF694RNA/@@download/ENCFF694RNA.bigWig",
  "tracks/heart_H3K27me3.bw", mode = "wb"
)
```

```{r load_peaks}
# Read all peak files
mesc_peaks <- list(
  p300     = rtracklayer::import("peaks/mesc_p300.bed.gz",     format = "narrowPeak"),
  H3K27ac  = rtracklayer::import("peaks/mesc_H3K27ac.bed.gz",  format = "narrowPeak"),
  H3K27me3 = rtracklayer::import("peaks/mesc_H3K27me3.bed.gz", format = "narrowPeak")
)

heart_peaks <- list(
  p300     = rtracklayer::import("peaks/heart_p300.bed.gz",     format = "narrowPeak"),
  H3K27ac  = rtracklayer::import("peaks/heart_H3K27ac.bed.gz",  format = "narrowPeak"),
  H3K27me3 = rtracklayer::import("peaks/heart_H3K27me3.bed.gz", format = "narrowPeak")
)

# Quick summary
lapply(mesc_peaks,  length)
lapply(heart_peaks, length)
```

---

# 2 – Question 1: Proportion of TSS overlapping each histone mark

## Building a TSS reference

```{r tss_reference}
# Extract protein-coding genes and derive ±500 bp TSS windows
pc_genes <- genes(ensdb, filter = GeneBiotypeFilter("protein_coding"))

# Fix chromosome naming to match ENCODE peaks (chr1, chr2, ...)
seqlevelsStyle(pc_genes) <- "UCSC"

# Keep only standard chromosomes present in the peaks
std_chroms <- seqlevels(mesc_peaks$H3K27ac)
pc_genes   <- keepSeqlevels(pc_genes, std_chroms, pruning.mode = "coarse")

tss <- promoters(pc_genes, upstream = 500, downstream = 500)
cat("Number of protein-coding TSS windows:", length(tss), "\n")
```

## Overlap proportions

```{r tss_overlaps}
tss_overlap <- function(tss, peaks_list) {
  has_ac   <- overlapsAny(tss, peaks_list$H3K27ac)
  has_me3  <- overlapsAny(tss, peaks_list$H3K27me3)

  c(
    H3K27ac_only       = mean( has_ac & !has_me3),
    H3K27me3_only      = mean(!has_ac &  has_me3),
    both               = mean( has_ac &  has_me3),
    H3K27ac            = mean( has_ac),             # any ac (incl. bivalent)
    H3K27me3           = mean( has_me3),
    neither            = mean(!has_ac & !has_me3)
  )
}

ov_mesc  <- tss_overlap(tss, mesc_peaks)
ov_heart <- tss_overlap(tss, heart_peaks)

# Display as a tidy table
tss_table <- data.frame(
  Category = names(ov_mesc),
  mESC     = round(ov_mesc  * 100, 1),
  Heart    = round(ov_heart * 100, 1)
)
print(tss_table, row.names = FALSE)
```

```{r tss_barplot, fig.width=7, fig.height=4}
# Reshape for ggplot
plot_df <- rbind(
  data.frame(Tissue = "mESC",  Category = names(ov_mesc),  Pct = ov_mesc  * 100),
  data.frame(Tissue = "Heart", Category = names(ov_heart), Pct = ov_heart * 100)
)
# Keep only the mutually exclusive categories for a stacked bar
plot_df <- subset(plot_df, Category %in% c("H3K27ac_only","H3K27me3_only","both","neither"))
plot_df$Category <- factor(plot_df$Category,
                            levels = c("H3K27ac_only","both","H3K27me3_only","neither"))

ggplot(plot_df, aes(x = Tissue, y = Pct, fill = Category)) +
  geom_col(width = 0.5) +
  scale_fill_manual(
    values = c(H3K27ac_only  = "#E69F00",
               both          = "#009E73",
               H3K27me3_only = "#56B4E9",
               neither       = "grey80"),
    labels = c("H3K27ac only (active)",
               "Both (bivalent)",
               "H3K27me3 only (repressed)",
               "Neither")
  ) +
  labs(y = "% of protein-coding TSS", x = NULL,
       title = "Chromatin state at protein-coding TSS",
       fill  = "Chromatin mark") +
  theme_bw(base_size = 13)
```

**Observations.**
In mESC, a sizeable fraction of TSS carry **H3K27me3 only** or **both marks** (bivalent
domains) – these are characteristic of developmental genes held in a poised state.
The bivalent category is much smaller in heart, where most TSS are either
unambiguously active (H3K27ac only) or fully repressed.
The proportion of TSS marked solely by H3K27ac is larger in heart, reflecting
the stable, committed transcriptional programme of a terminally differentiated
tissue.

---

# 3 – Question 2: Relationship between p300, H3K27ac and H3K27me3

## Pairwise peak overlaps within each tissue

```{r peak_overlaps}
regionOverlaps(mesc_peaks,  title = "mESC – peak overlaps")
regionOverlaps(heart_peaks, title = "Heart – peak overlaps")
```

## Signal heatmaps centred on p300 peaks

We visualise all three signals around p300-bound sites, which mark active
enhancers, to see whether H3K27ac co-localises and whether H3K27me3 is excluded.

```{r signal_heatmap_mesc, fig.width=8, fig.height=6}
bw_mesc <- list(
  p300     = "tracks/mesc_p300.bw",
  H3K27ac  = "tracks/mesc_H3K27ac.bw",
  H3K27me3 = "tracks/mesc_H3K27me3.bw"
)

sm_mesc <- signal2Matrix(bw_mesc, regions = mesc_peaks$p300, extend = 2000, w = 50)
sm_mesc <- renormalizeSignalMatrices(sm_mesc, method = "top", trim = 0.05)

plotEnrichedHeatmaps(
  sm_mesc, trim = 0.95, multiScale = TRUE,
  colors = list("darkred", "#E69F00", "#56B4E9"),
  row_title = "mESC p300 peaks"
)
```

```{r signal_heatmap_heart, fig.width=8, fig.height=6}
bw_heart <- list(
  p300     = "tracks/heart_p300.bw",
  H3K27ac  = "tracks/heart_H3K27ac.bw",
  H3K27me3 = "tracks/heart_H3K27me3.bw"
)

sm_heart <- signal2Matrix(bw_heart, regions = heart_peaks$p300, extend = 2000, w = 50)
sm_heart <- renormalizeSignalMatrices(sm_heart, method = "top", trim = 0.05)

plotEnrichedHeatmaps(
  sm_heart, trim = 0.95, multiScale = TRUE,
  colors = list("darkred", "#E69F00", "#56B4E9"),
  row_title = "Heart p300 peaks"
)
```

**Observations.**
In both tissues, p300 co-localises tightly with **H3K27ac**, confirming their
shared role in marking active enhancers. H3K27me3 signal is consistently
depleted at p300-bound sites, as expected given that these two states are
mutually exclusive: p300/H3K27ac mark actively used regulatory elements, while
H3K27me3 marks Polycomb-repressed chromatin.
The relationship is similar in both tissues, but the number and location of
p300/H3K27ac co-marked enhancers differs substantially between mESC and heart,
reflecting tissue-specific enhancer usage.

---

# 4 – Question 3: Regulatory elements active in mESC vs heart

## Tissue-specific enhancers

We define "mESC-specific" and "heart-specific" regulatory elements as p300
peaks that do not overlap any p300 peak in the other tissue.

```{r tissue_specific}
mesc_only  <- mesc_peaks$p300[!overlapsAny(mesc_peaks$p300,  heart_peaks$p300)]
heart_only <- heart_peaks$p300[!overlapsAny(heart_peaks$p300, mesc_peaks$p300)]
shared     <- mesc_peaks$p300[ overlapsAny(mesc_peaks$p300,  heart_peaks$p300)]

cat(sprintf(
  "mESC-specific p300 sites: %d\nHeart-specific p300 sites: %d\nShared: %d\n",
  length(mesc_only), length(heart_only), length(shared)
))
```

## Signal at mESC-specific sites in both tissues

```{r mesc_specific_signal, fig.width=8, fig.height=6}
bw_both <- c(
  mesc_p300    = "tracks/mesc_p300.bw",
  mesc_H3K27ac = "tracks/mesc_H3K27ac.bw",
  heart_p300   = "tracks/heart_p300.bw",
  heart_H3K27ac= "tracks/heart_H3K27ac.bw"
)

sm_mesc_only <- signal2Matrix(bw_both, regions = mesc_only, extend = 2000, w = 50)
sm_mesc_only <- renormalizeSignalMatrices(sm_mesc_only, method = "top", trim = 0.05)

plotEnrichedHeatmaps(
  sm_mesc_only, trim = 0.95, multiScale = TRUE,
  row_title = "mESC-specific p300 sites"
)
```

## Signal at heart-specific sites in both tissues

```{r heart_specific_signal, fig.width=8, fig.height=6}
sm_heart_only <- signal2Matrix(bw_both, regions = heart_only, extend = 2000, w = 50)
sm_heart_only <- renormalizeSignalMatrices(sm_heart_only, method = "top", trim = 0.05)

plotEnrichedHeatmaps(
  sm_heart_only, trim = 0.95, multiScale = TRUE,
  row_title = "Heart-specific p300 sites"
)
```

## UpSet plot – full overlap structure

```{r upset, fig.width=7, fig.height=5}
all_peaks <- list(
  mesc_p300    = mesc_peaks$p300,
  mesc_H3K27ac = mesc_peaks$H3K27ac,
  heart_p300   = heart_peaks$p300,
  heart_H3K27ac= heart_peaks$H3K27ac
)

comb_mat <- make_comb_mat(
  regionsToUpset(all_peaks, nsets = length(all_peaks))
)
ComplexHeatmap::UpSet(comb_mat,
  comb_order = order(comb_size(comb_mat), decreasing = TRUE),
  top_annotation = upset_top_annotation(comb_mat, add_numbers = TRUE)
)
```

**Observations.**
The majority of p300/H3K27ac-marked enhancers are **tissue-specific**: elements
active in mESC are largely inactive (no p300 binding, low H3K27ac) in heart,
and vice versa. This demonstrates that cell identity is encoded to a large
extent at the level of enhancer activity, not just at promoters.
Shared elements tend to correspond to housekeeping enhancers or broadly
active regulatory elements.
The near-complete loss of mESC-specific enhancer activity in heart (and
vice versa) is consistent with enhancer switching during differentiation,
where lineage-determining transcription factors activate new sets of enhancers
while previously active ones are silenced, often accompanied by gain of
H3K27me3 at disused regulatory elements.

---

# Session info

```{r session_info}
sessionInfo()
```
