---
title: "Week 09 Assignment"
author: "Thomas"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(GenomicRanges)
  library(rtracklayer)
  library(SummarizedExperiment)
  library(S4Vectors)
  library(epiwraps)
  library(ggplot2)
  library(rGREAT)
})

set.seed(123)
```

## Part 1: Download and prepare the data

```{r download}
data_dir <- if (basename(getwd()) == "week09") "." else "week09"
zip_file <- file.path(data_dir, "clustering_assignment.zip")

if (!file.exists(zip_file)) {
  options(timeout = 6000)
  download.file(
    "https://ethz-ins.org/content/clustering_assignment.zip",
    destfile = zip_file,
    mode = "wb"
  )
}

if (length(list.files(data_dir, pattern = "\\.(bed|bw|bigWig)$", ignore.case = TRUE)) == 0) {
  unzip(zip_file, exdir = data_dir)
}

bed_files <- list.files(data_dir, pattern = "\\.bed$", full.names = TRUE, ignore.case = TRUE)
track_files <- list.files(data_dir, pattern = "\\.(bw|bigWig)$", full.names = TRUE, ignore.case = TRUE)
names(track_files) <- tools::file_path_sans_ext(basename(track_files))

bed_files
track_files
```

```{r regions}
peaks <- lapply(bed_files, rtracklayer::import.bed)
names(peaks) <- tools::file_path_sans_ext(basename(bed_files))

high_confidence_peaks <- lapply(peaks, function(x) {
  if ("score" %in% colnames(S4Vectors::mcols(x))) {
    x[x$score > 800]
  } else {
    x
  }
})

regions <- GenomicRanges::reduce(unlist(GenomicRanges::GRangesList(high_confidence_peaks)))
regions <- GenomicRanges::sort(regions)

lengths(high_confidence_peaks)
length(regions)
```

## Part 2: Cluster the binding profiles

```{r signal-matrix, fig.width=9, fig.height=7}
ese <- epiwraps::signal2Matrix(track_files, regions, extend = 2000)

epiwraps::plotEnrichedHeatmaps(
  ese,
  cluster_rows = TRUE,
  show_row_dend = TRUE,
  colors = c("white", "darkred")
)
```

```{r choose-k, fig.width=7, fig.height=4}
set.seed(123)
cl_range <- epiwraps::clusterSignalMatrices(ese, k = 2:12)

ggplot(cl_range$varExplained, aes(k, varExplained)) +
  geom_line() +
  geom_point() +
  labs(
    x = "Number of clusters",
    y = "Variance explained",
    title = "Choosing the number of clusters"
  ) +
  theme_classic()
```

The variance curve decreased gradually rather than showing a sharp elbow, so there was no single obvious optimal value of `k`. I therefore chose `k = 6` as a compromise: it separates the main binding patterns visible in the heatmap while keeping enough regions in each cluster for downstream enrichment analysis.

```{r cluster-profiles, fig.width=9, fig.height=7}
set.seed(123)
cl <- epiwraps::clusterSignalMatrices(ese, k = 6, scaleRows = TRUE)
rowData(ese)$cluster <- factor(cl)

table(rowData(ese)$cluster)

cluster_colors <- c(
  "1" = "#D55E00",
  "2" = "#0072B2",
  "3" = "#009E73",
  "4" = "#CC79A7",
  "5" = "#E69F00",
  "6" = "#56B4E9"
)

epiwraps::plotEnrichedHeatmaps(
  ese,
  row_split = "cluster",
  scale_rows = "global",
  mean_color = cluster_colors,
  colors = c("white", "darkred")
)
```

```{r cluster-averages, fig.width=9, fig.height=6}
cluster_means <- epiwraps::meltSignals(ese, splitBy = rowData(ese)$cluster)

ggplot(cluster_means, aes(position, mean, colour = sample)) +
  geom_line(linewidth = 1) +
  facet_wrap(~split) +
  labs(
    x = "Position relative to peak center",
    y = "Mean ChIP-seq signal",
    colour = "Factor",
    title = "Average binding profiles by cluster"
  ) +
  theme_classic()
```

## Part 3: Gene Ontology enrichment

```{r split-regions}
split_regions <- split(rowRanges(ese), rowData(ese)$cluster)
lengths(split_regions)
```

```{r great-all-clusters, message=FALSE, warning=FALSE}
enrichment_tables <- lapply(names(split_regions), function(cluster_id) {
  res <- rGREAT::great(
    split_regions[[cluster_id]],
    gene_sets = "GO:BP",
    tss_source = "hg38",
    background = regions,
    cores = 2
  )

  tab <- rGREAT::getEnrichmentTables(res)
  tab[order(tab$p_adjust, tab$p_adjust_hyper), ]
})

names(enrichment_tables) <- names(split_regions)

enrichment_summary <- do.call(rbind, lapply(names(enrichment_tables), function(cluster_id) {
  tab <- enrichment_tables[[cluster_id]]

  data.frame(
    cluster = cluster_id,
    n_regions = length(split_regions[[cluster_id]]),
    top_term = tab$description[1],
    fold_enrichment = tab$fold_enrichment[1],
    p_adjust = tab$p_adjust[1],
    p_adjust_hyper = tab$p_adjust_hyper[1]
  )
}))

enrichment_summary
```

```{r select-enrichment-cluster}
cluster_for_enrichment <- as.character(
  enrichment_summary$cluster[which.min(enrichment_summary$p_adjust)]
)

bp <- enrichment_tables[[cluster_for_enrichment]]
bp <- bp[order(bp$p_adjust, bp$p_adjust_hyper), ]

cluster_for_enrichment
length(split_regions[[cluster_for_enrichment]])
head(bp, 10)
```

```{r enrichment-plot, fig.width=9, fig.height=6}
top_bp <- head(bp, 15)

ggplot(
  top_bp,
  aes(fold_enrichment, reorder(description, p_adjust),
      size = observed_region_hits, color = -log10(p_adjust))
) +
  geom_point() +
  scale_color_viridis_c() +
  labs(
    x = "Fold enrichment",
    y = "GO biological process",
    size = "Region hits",
    color = "-log10 adjusted P",
    title = paste("GO enrichment for cluster", cluster_for_enrichment)
  ) +
  theme_classic()
```

## Results

```{r results-paragraph, results='asis'}
cluster_sizes <- table(rowData(ese)$cluster)
largest_cluster <- names(cluster_sizes)[which.max(cluster_sizes)]
top_terms <- paste(head(bp$description, 3), collapse = ", ")

cat(
  "After filtering the CREB-family peak sets for high-confidence peaks, I obtained ",
  length(regions),
  " non-redundant chromosome 1 regions. I clustered the ChIP-seq signal over these regions into six groups, with cluster ",
  largest_cluster,
  " being the largest (",
  max(cluster_sizes),
  " regions). The average profile plot showed both protein-specific groups and groups where multiple CREB-family factors were enriched at the same regions, indicating that the three factors do not occupy all high-confidence regions in the same way. Cluster 1 was mainly enriched for CREB1, cluster 2 showed CREB1 and CREB3L1 together, cluster 3 was mainly enriched for CREB3, cluster 4 showed all three factors with the strongest signal for CREB3L1 followed by CREB3 and CREB1, cluster 5 showed CREB3L1 followed by CREB1 and weaker CREB3 signal, and cluster 6 was also CREB1-enriched but with a different balance of CREB3 and CREB3L1 compared with cluster 1. I compared GO Biological Process enrichment across all six clusters and selected cluster ",
  cluster_for_enrichment,
  " for detailed visualization because it had the smallest adjusted enrichment P-value. The top enriched GO Biological Process terms in this cluster included ",
  top_terms,
  ". Because cluster ",
  cluster_for_enrichment,
  " showed stronger CREB3 signal than the other CREB-family factors, these results suggest that CREB3-bound regions may be associated with genes involved in cell communication and signaling-related processes.",
  sep = ""
)
```
